---
title: "1st Foray into SQL in R"
subtitle: "Integration of SQL and R to produce visualisations"
author: "Rohit" 
output: 
  html_document:
    theme: cosmo
    toc: true            # Generates a Table of Contents
    toc_float: true      # Makes the TOC float on the left side as you scroll
    highlight: tango     # Makes your SQL/R syntax highlighting look beautiful
    code_folding: hide   # Keeps the code panels closed on first view
    self_contained: true
---

```{r color_them, include=FALSE}
library(ggplot2)

# 1. SHARED DESIGN SYSTEM (Run this once at the top of your visualization section)
brand_colors <- c("#4A6572", "#34495E", "#1A252C", "#E07A5F")

professional_theme <- theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 13, color = "#1A252C", margin = margin(b = 4)),
    plot.subtitle = element_text(size = 9.5, color = "#627D98", margin = margin(b = 15)),
    axis.title = element_text(face = "bold", size = 10, color = "#334E68"),
    axis.text = element_text(size = 9, color = "#486581"),
    panel.grid.major = element_line(color = "#F0F4F8", linewidth = 0.5),
    panel.grid.minor = element_blank(),
    legend.position = "top",
    legend.justification = "left",
    plot.margin = margin(15, 15, 15, 15),
    
    # NEW: Professional styling for the bottom source text
    plot.caption = element_text(face = "italic", size = 8, color = "#94A3B8", margin = margin(t = 15)),
    plot.caption.position = "plot" # Left-aligns the caption with the whole plot canvas, not just the grid
  )
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,        # Shows your SQL and R code in the final report (great for portfolios)
  results = "hide",   # HIDES all text data table outputs globally!
  message = FALSE,    # Hides annoying package loading messages
  warning = FALSE,    # Hides warning messages (like ggplot deprecated guides)
  comment = NA,       # Removes the default "##" prefix from R console outputs
  fig.align = "center" # Centers all your ggplot graphs automatically
)

# 1. Load libraries
library(DBI)
library(RMariaDB)  # Swap for RPostgreSQL if using Postgres
library(tidyverse)
library(car)       # For advanced statistical testing if needed
library(tidytext)

# 2. Establish connection to your local database server
# Make sure your local MySQL/Postgres server is running!
con <- dbConnect(RMariaDB::MariaDB(),
                 user = "root",
                 password = "password",
                 host = "host",
                 port = port)

# 3. Read and execute your Parks & Rec SQL file to build the database and tables
sql_file_path <- "./Beginner - Parks_and_Rec_Create_db.sql"

if (file.exists(sql_file_path)) {
  sql_script <- readLines(sql_file_path)
  # Join lines and split by semicolon to run statements individually
  sql_commands <- unlist(strsplit(paste(sql_script, collapse = "\n"), ";"))
  
  for (cmd in sql_commands) {
    clean_cmd <- trimws(cmd)
    if (clean_cmd != "") {
      dbExecute(con, clean_cmd)
    }
  }
  
  # 4. Disconnect from bare server and reconnect directly to the new database
  dbDisconnect(con)
  con <- dbConnect(RMariaDB::MariaDB(),
                   user = "root",
                   password = "password",
                   host = "host",
                   port = port,
                   dbname = "parks_and_recreation")
  
  # 5. Set global connection for all future SQL chunks
  knitr::opts_chunk$set(connection = "con")
  
} else {
  stop("SQL file not found! Double check your file path string.")
}
```

```{r test_connection, include=FALSE}
# This will print a list of all tables inside your database
dbListTables(con)
```

## Introduction

This notebook marks the beginning of my end-to-end data analysis portfolio. By combining the data manipulation power of **SQL** with the statistical and visual capabilities of **R**, I am building a reproducible environment to re-master advanced database concepts and translate raw data into compelling data stories.

The data used throughout this project is based on the *Parks and Recreation* and the *2020-2023 Company Layoffs* datasets, covering foundational queries up to advanced window functions, CTEs, and optimization techniques.

------------------------------------------------------------------------

## Foundational Queries & Filtering

### 1. SELECT statements

Pulling all columns and rows from the `employee_demographics` table to inspect our baseline population.

```{sql SELECT,connection=con, echo = TRUE}
SELECT 
  employee_id,
  first_name,
  last_name,
  age,
  (age+10) AS age_in_a_decade
FROM employee_demographics;
```

```{sql DISTINCT, connection=con, echo = TRUE}
SELECT DISTINCT gender
FROM employee_demographics;

```

```{sql DEM, connection=con, echo = TRUE}
SELECT *
FROM employee_demographics;

```

```{sql SAL, connection=con, echo = TRUE}
SELECT *
FROM employee_salary;

```

### 2. Aggregating Data with GROUP BY

The `GROUP BY` clause collapses rows that have the same values into summary rows. Finding out how many employees belong to each gender and calculate their average age.

When using `GROUP BY`, remember any column in your `SELECT` statement that is not part of an aggregate function (like `COUNT`, `AVG`, `MIN`, `MAX`) **must** be listed in the `GROUP BY` clause.

```{sql GROUP BY, connection=con, echo = TRUE}
SELECT 
    gender,
    COUNT(employee_id) AS total_employees,
    AVG(age) AS average_age,
    MIN(age) AS youngest_employee,
    MAX(age) AS oldest_employee
FROM employee_demographics
GROUP BY gender;
```

#### 2.1 HAVING vs WHERE

WHERE filters rows before any grouping happens.

HAVING filters rows after the GROUP BY aggregation has been calculated.

```{sql HAVING vs WHERE, connection=con,echo = TRUE}
SELECT dept_id,
    COUNT(employee_id) AS staff_count,
    ROUND(AVG(salary), 2) AS average_salary
FROM employee_salary
GROUP BY dept_id
HAVING AVG(salary) > 70000;
```

### 3. Sorting and Limiting Results using ORDER BY and LIMIT

```{sql ORDER BY and LIMIT, connection=con,echo = TRUE}
SELECT 
    employee_id,
    first_name,
    last_name,
    salary
FROM employee_salary
ORDER BY salary DESC
LIMIT 3;
```

### 4. JOINS

Joins allow linking tables together using a common column.

##### 4.1 Inner Join

An Inner Join only returns records where the `employee_id` matches in both tables.

```{sql Inner Join, connection=con,echo = TRUE}
SELECT 
    dem.employee_id, 
    dem.first_name, 
    dem.last_name, 
    sal.occupation, 
    sal.salary
FROM employee_demographics dem
INNER JOIN employee_salary sal
    ON dem.employee_id = sal.employee_id;
```

##### 4.2. Left Join

A Left Join keeps all rows from the left table (`employee_demographics`), even if they don't have matching salary data.

```{sql Left Join, connection=con,echo = TRUE}
SELECT 
    dem.employee_id, 
    dem.first_name, 
    dem.last_name, 
    sal.occupation, 
    sal.salary
FROM employee_demographics dem
LEFT JOIN employee_salary sal
    ON dem.employee_id = sal.employee_id;
```

##### 4.3 Right Join

Keeping all rows from `employee_salary` and matching demographic data.

```{sql Right Join, connection=con,echo = TRUE}
SELECT 
    sal.employee_id, 
    sal.first_name, 
    sal.last_name, 
    dem.age, 
    dem.gender
FROM employee_salary sal
LEFT JOIN employee_demographics dem
    ON sal.employee_id = dem.employee_id;
```

##### 4.4 Self Join

Ties a table to itself.

```{sql SELF JOIN, connection=con,echo = TRUE}
SELECT 
    emp1.employee_id    AS emp_santa, 
    emp2.employee_id    AS emp_name, 
    CONCAT(emp1.first_name, ' ', emp1.last_name) AS name_santa,
    CONCAT(emp2.first_name, ' ', emp2.last_name) AS name_emp
FROM employee_salary AS emp1
LEFT JOIN employee_salary AS emp2
    ON emp1.employee_id + 1 = emp2.employee_id;
```

##### 4.5 Joining Multiple Tables

```{sql, connection=con,echo = TRUE}
# --- JOINING MULTIPLE TABLES together ---
select dem.first_name, dem.last_name, occupation,department_name
from employee_demographics AS dem
inner join employee_salary AS sal
	ON dem.employee_id = sal.employee_id
inner join parks_departments as pd
	ON sal.dept_id = pd.department_id
;
```

### 5. Unions

Combining ROWS of data ACROSS tables`DISTINCT` is for only unique values while `ALL` is used to show all values.

```{sql UNIONS, connection=con,echo = TRUE}
SELECT first_name, last_name
FROM employee_demographics as dem
UNION ALL
SELECT first_name, last_name
FROM employee_salary;
```

```{sql Categorizing Employees, connection=con,echo = TRUE}
SELECT first_name, last_name, 'Old Man' As Label
FROM employee_demographics
WHERE age > 40 AND gender = "Male"
UNION
SELECT first_name, last_name, 'Old Lady' As Label
FROM employee_demographics
WHERE age > 40 AND gender = "Female"
UNION
SELECT first_name, last_name, 'Highly Paid Employee' As Label
FROM employee_salary
WHERE salary > 70000
ORDER BY first_name, last_name;
```

Goal: Create a highly targeted list of employees based on distinct business rules.

```{sql Identifying the Veterans/High Earners, connection=con,echo = TRUE}
SELECT 
    demo.first_name, 
    demo.last_name, 
    'Highly Paid Veteran' AS employee_status
FROM employee_demographics demo
JOIN employee_salary sal
    ON demo.employee_id = sal.employee_id
WHERE demo.age > 55 AND sal.salary > 40000

UNION

SELECT 
    demo.first_name, 
    demo.last_name,
    'Highly Paid Rising Star' AS employee_status
FROM employee_demographics demo
JOIN employee_salary sal
    ON demo.employee_id = sal.employee_id
WHERE demo.age < 40 AND sal.salary > 40000

ORDER BY last_name ASC; 
-- Note: The ORDER BY clause goes at the very end of the entire statement,
--       and it sorts the final combined result set!
```

### 6. TRIM, STRING, SUBSTRING and REPLACE

REAL WORLD CASE - Phone `#` length

```{sql LENGTH, connection=con,echo = TRUE}
select first_name, length(first_name)
FROM employee_demographics
ORDER BY 2;
```

```{sql UPPER/LOWER, connection=con, include=FALSE}
-- select upper('sky');
-- select lower('SKY');
```

```{sql UPPER, connection=con}
select first_name, upper(first_name)
FROM employee_demographics
;
```

TRIM --- Removes Whitespace on LEFT or RIGHT of value

##### 6.1 SUBSTRINGS

RIGHT and LEFT SUBSTRINGS

```{sql LEFT, connection=con,echo = TRUE}
select first_name, left(first_name,4)
from employee_demographics;
```

Substrings

Can get middle of words or dates SUBSTRING(string_expression, start_position, length)

```{sql SUBSTRING, connection=con,echo = TRUE}
select first_name, 
left(first_name,4),
right(first_name,4),
substring(first_name,3,2),
birth_date,
substring(birth_date,6,2) as birth_month
from employee_demographics;
```

##### 6.2 REPLACE --- Replaces characters with preferred ones ---

```{sql REPLACE, connection=con,echo = TRUE}
select first_name, replace(first_name, 'a','z')
from employee_demographics;
```

```{sql CONCAT2, connection=con,echo = TRUE}
# LOCATE 
# CONCAT --- Combine fields
select first_name, last_name,
concat(first_name,' ', last_name) As Full_Name
from employee_demographics;
```

### 7. CASE Statements

Goal: Create dynamic columns based on multiple conditional rules.

Note: Every CASE statement must start with `CASE`, contain at least -- one `WHEN/THEN`, an optional `ELSE`, and must close with `END`.

```{sql CASE, connection=con,echo = TRUE}
SELECT 
    first_name,
    last_name,
    salary,
    dept_id,
    
    -- Example 1: Creating a Text Category (Salary Bands)
    CASE 
        WHEN salary >= 60000 THEN 'High Earner'
        WHEN salary BETWEEN 45000 AND 69999 THEN 'Mid-Tier'
        ELSE 'Entry Level'
    END AS salary_bracket,
    
    -- Example 2: Performing Math (Dynamic Bonus Structure)
    -- If they work in Parks (dept_id = 1), they get a 10% bonus.
    -- If they work in Finance (dept_id = 4), they get a 5% bonus.
    -- Everyone else gets a 2% bonus.
    CASE
        WHEN dept_id = 1 THEN salary * 0.10
        WHEN dept_id = 4 THEN salary * 0.05
        ELSE salary * 0.02
    END AS holiday_bonus

FROM employee_salary
ORDER BY holiday_bonus DESC;
```

### 8. Subqueries

```{sql Subqueries, connection=con,echo = TRUE}
select * 
from employee_demographics
where employee_id in 
				(select employee_id
					from employee_salary
                    where dept_id = 1)
;
```

```{sql Subqueries2,connection=con,echo = FALSE}
select first_name, salary, 
(select avg(salary)
from employee_salary) As Avg_Salary
from employee_salary;
```

```{sql Gender Stats,connection=con,echo = TRUE}
select gender, avg(age), max(age),min(age),count(age)
from employee_demographics
group by gender;
```

```{sql Subqueries3,connection=con,echo = FALSE}
select first_name, salary, 
(select avg(salary)
from employee_salary) As Avg_Salary
from employee_salary;
```

```{sql Gender Table,connection=con,echo = TRUE}
select avg(max_age)
from (select 
	gender, 
	avg(age) as avg_age, 
	max(age) as max_age,
	min(age) as min_age,
	count(age)
from employee_demographics
group by gender) As Aggregated_Table;
```

```{sql Subqueries4,connection=con,echo = TRUE}
select first_name, salary, 
(select avg(salary)
from employee_salary) As Avg_Salary
from employee_salary;
```

### 9. Window Functions

```{sql Window Functions,connection=con,echo = TRUE}
SELECT gender, avg(salary) as Avg_Salary
FROM employee_demographics as dem
JOIN employee_salary as sal
	ON dem.employee_id=sal.employee_id
group by gender;
```

```{sql Window Functions2,connection=con,echo = TRUE}
SELECT concat(dem.first_name,' ',dem.last_name) AS Name,gender, avg(salary) OVER(partition by gender) As Avg_Salary
FROM employee_demographics as dem
JOIN employee_salary as sal
	ON dem.employee_id=sal.employee_id;
```

##### 9.1 Rolling Totals

Start at a value and add rows below based on partition

```{sql Rolling Total,connection=con,echo = TRUE}
SELECT concat(dem.first_name,' ',dem.last_name) AS Name,gender,salary, sum(salary) OVER(partition by gender ORDER BY dem.employee_id) As RollingTotal
FROM employee_demographics as dem
JOIN employee_salary as sal
	ON dem.employee_id=sal.employee_id;
```

###### 9.2 Special Window Functions

```{sql Special Window Functions,connection=con,echo = TRUE}
SELECT gender, avg(salary) as Avg_Salary
FROM employee_demographics as dem
JOIN employee_salary as sal
	ON dem.employee_id=sal.employee_id
group by gender;
```

##### 9.3 Ranking Functions

1.  **ROW_NUMBER()**: The Strict Counter It assigns a strict, sequential number (1, 2, 3, 4...) to every single row.

    1.  Crucial Rule: It NEVER allows ties. If two people make the exact same salary, it will arbitrarily give one a '2' and the other a '3'.

2.  **RANK()**: The Olympic Medalist It allows ties. If two people have the exact same salary, they get the SAME rank.

    1.  Crucial Rule: It SKIPS numbers after a tie.

    2.  Example: If there's a tie for 2nd place, the output is: 1, 2, 2, 4 (it skips 3),

3.  **DENSE_RANK()**: The Dense/No-Skip Ranker Like RANK(), it allows ties and gives identical salaries the same rank number.

    1.  Crucial Rule: It NEVER skips numbers. It keeps the ranking compact ("dense").

    2.  Example: If there's a tie for 2nd place, the output is: 1, 2, 2, 3 (next person is 3).

```{sql Ranking,connection=con,echo = TRUE}
SELECT  dem.employee_id,concat(dem.first_name,' ',dem.last_name) AS Name,gender,salary,
ROW_NUMBER() OVER(PARTITION BY gender ORDER BY salary DESC) As row_num,
RANK() OVER(PARTITION BY gender ORDER BY salary DESC) As rank_num,
DENSE_RANK() OVER(PARTITION BY gender ORDER BY salary DESC) As dense_rank_num	
FROM employee_demographics as dem
JOIN employee_salary as sal
	ON dem.employee_id=sal.employee_id;
```

### 10. Common Table Expression (CTE)

A temporary, named result set which uses the `WITH` function.

```{sql Creating a CTE,connection=con,echo = TRUE}
WITH CTE_Example (Gender, Avg_Sal,Max_Sal,Min_Sal,Count_Sal)As
(
SELECT gender, avg(salary), max(salary),min(salary),count(salary)
from employee_demographics dem
join employee_salary sal
	on dem.employee_id=sal.employee_id
group by gender
)
SELECT *
FROM CTE_Example
;
```

```{sql Joining Tables using CTEs,connection=con,echo = TRUE}
WITH CTE_Example As
(
SELECT employee_id,gender,birth_date
from employee_demographics
where birth_date > '1985-01-01'
),
CTE_Example2 As
(
select employee_id,salary
from employee_salary
where salary > 50000
)
SELECT *
FROM CTE_Example
JOIN CTE_Example2
	ON CTE_Example.employee_id=CTE_Example2.employee_id
;
```

### 11. Stored Procedures

Creating procedures that you run

```{sql Stored Procedures,connection=con,echo = TRUE}
select *
from employee_salary;
```

#### 11.1 PARAMETERS -- Passed as an INPUT into a Stored Procedure

```{R Creating a Stored Procedure,connection=con,echo = TRUE}
create_proc_sql <- "
CREATE PROCEDURE salary_finder(p_employee_id INT)
BEGIN
    SELECT salary
    FROM employee_salary
    WHERE employee_id = p_employee_id;
END;
"

# Use dbExecute to create the procedure
dbExecute(con, create_proc_sql)
```

```{r, ,echo = TRUE}
# Call the procedure for employee_id = 7
result <- dbExecute(con, "CALL salary_finder(6);")

# View the result
print(result)
```

### 12. Triggers and Events

BLOCK of CODE that automatically executes when an `EVENT` occurs in a table `AFTER` INSERT - after data is added `BEFORE` INSERT - if data is deleted from the table

```{sql Employee Demo,connection=con,echo = TRUE}
SELECT *
from employee_demographics;
```

```{sql Employee Salaries,connection=con,echo = TRUE}
SELECT *
from employee_salary;
```

##### 12.1 Define the Trigger SQL inside an R multiline string.

```{r Trigger SQL,echo = TRUE}
dbExecute(con, "DROP TRIGGER IF EXISTS employee_insert;")

trigger_blueprint <- "
CREATE TRIGGER employee_insert
    AFTER INSERT ON employee_salary
    FOR EACH ROW
BEGIN
    INSERT INTO employee_demographics (employee_id, first_name, last_name)
    VALUES (NEW.employee_id, NEW.first_name, NEW.last_name);
END;
"

# ==========================================
# STEP 3: Compile into the Database
# ==========================================
dbExecute(con, trigger_blueprint)
```

```{r Executing a Trigger,echo = TRUE}
# Insert data into the primary table only
dbExecute(con, "
  INSERT INTO employee_salary (employee_id, first_name, last_name, salary) 
  VALUES (15, 'Ron', 'Swanson', 80000);
")

# Query the second table to verify the side-effect worked perfectly
side_effect_check <- dbExecute(con, "SELECT * FROM employee_demographics WHERE employee_id = 15;")
print(side_effect_check)
```

```{r EVENT_SCHEDULER, include=FALSE, echo=FALSE}
dbExecute(con, "SET GLOBAL event_scheduler = ON;")
print("Success")
```

### 13. Events

Takes place when SCHEDULED

```{r Events,echo = TRUE}
# 2. Define the Event logic as an R string
event_sql <- "
CREATE EVENT delete_retirees
ON SCHEDULE EVERY 2 SECOND         -- 1. The Clock: Tells the DB how often to wake up
DO 
BEGIN                              -- 2. The Task: What happens when it wakes up
    DELETE FROM employee_demographics
    WHERE age >= 60;
END;
"

# 3. Compile the event into the database
dbExecute(con, event_sql)

print("Time-based Event successfully scheduled!")
```

```{r ViewDem,echo = FALSE}
# View demographics before the event catches older employees
print(dbExecute(con, "SELECT * FROM employee_demographics;"))
```

```{r Check_Auto, include=FALSE}
# Pause R for 3 seconds to let the database event fire
#Sys.sleep(3)

# Check the table again to see the automation side-effect
print(dbExecute(con, "SELECT * FROM employee_demographics;"))

# To PAUSE the event without deleting it:
dbExecute(con, "ALTER EVENT delete_retirees DISABLE;")

# To RESUME the event:
dbExecute(con, "ALTER EVENT delete_retirees ENABLE;")

# To completely DELETE the event:
dbExecute(con, "DROP EVENT IF EXISTS delete_retirees;")
```

## PROJECT: Data Cleaning

Data Cleaning the *2020-2023 Company Layoffs data* for Analysis

> Step 1: De-duplicate values
>
> Step 2: Standardize data
>
> Step 3: Null or Blank values
>
> Step 4: Remove Any Columns or Rows

```{r Importing a CSV,echo = TRUE}
dbDisconnect(con)
# 1. Create an in-memory SQLite database connection
con <- dbConnect(RSQLite::SQLite(), ":memory:")
knitr::opts_chunk$set(connection = "con")
  
# 2. Read your local CSV file into R
layoffs_raw <- read_csv("layoffs.csv")

# 3. Write the data frame into your SQL connection as a table named "layoffs"
dbWriteTable(con, "layoffs", layoffs_raw, overwrite = TRUE)
```

```{sql DATA CLEANING,connection=con,echo = TRUE}
-- View all records from the initial raw layoffs table
SELECT * FROM layoffs LIMIT 10;
```

### Step 1: Creating a Duplicate Table to preserve raw data

```{sql Creating a Duplicate Table,connection=con,echo = TRUE}
-- Create and populate the first staging table from the raw data
CREATE TABLE layoffs_staging AS
SELECT * FROM layoffs;
```

```{sql VerifyMigrationofData,connection=con}
-- Verify the migration
SELECT * FROM layoffs_staging LIMIT 5;
```

### Step 2: Check for duplicate values

```{sql Check for duplicates where row_num > 1,connection=con,echo = TRUE}
WITH duplicate_cte AS (
    SELECT *,
    ROW_NUMBER() OVER(
        PARTITION BY company, location, industry, total_laid_off, 
                     percentage_laid_off, date, stage, country, funds_raised_millions
    ) AS row_num
    FROM layoffs_staging
)
SELECT * FROM duplicate_cte 
WHERE row_num > 1;
```

Looking at Casper company

```{sql Pick a Company,connection=con,echo = TRUE}
SELECT *
FROM layoffs_staging
WHERE company = 'Casper';
```

```{sql Layoffs,connection=con,echo = TRUE}
SELECT *,
ROW_NUMBER() OVER(
PARTITION BY company,location,industry, total_laid_off, percentage_laid_off, `date`,stage,country,funds_raised_millions) AS ROW_NUM
FROM layoffs_staging;
```

```{sql Create Duplicate Table,connection=con,echo = TRUE}
CREATE TABLE `layoffs_staging2` (
  `company` text,
  `location` text,
  `industry` text,
  `total_laid_off` int DEFAULT NULL,
  `percentage_laid_off` text,
  `date` text,
  `stage` text,
  `country` text,
  `funds_raised_millions` int DEFAULT NULL,
  `row_num` int
);
```

```{sql DataOverview,connection=con,echo = TRUE}
SELECT *
FROM layoffs_staging2;
```

```{sql Duplicate Staging Table,connection=con,echo = TRUE}
INSERT INTO layoffs_staging2
SELECT *,
ROW_NUMBER() OVER(
PARTITION BY company,location,industry, total_laid_off, percentage_laid_off, `date`,stage,country,funds_raised_millions) AS ROW_NUM
FROM layoffs_staging;
```

```{sql Check for Duplicates,connection=con,echo = TRUE}
SELECT *
FROM layoffs_staging2
WHERE row_num > 1;
```

```{sql Delete nrows > 1,connection=con,echo = TRUE}
DELETE
FROM layoffs_staging2
WHERE row_num > 1;
```

```{sql Check for Duplicates Again,connection=con}
SELECT *
FROM layoffs_staging2;
```

### Step 3: Standardizing data

```{sql Trimming Company Data,connection=con,echo = TRUE}
SELECT company, TRIM(company)
FROM layoffs_staging2;
```

```{sql Update_Trimming Company Data,connection=con,echo = TRUE}
UPDATE layoffs_staging2
SET company = TRIM(company);
```

```{sql Check_For_Crypto,connection=con,echo = TRUE}
SELECT *
FROM layoffs_staging2
WHERE industry LIKE 'Crypto%';
```

```{sql Update_Check_For_Crypto,connection=con,echo = TRUE}
UPDATE layoffs_staging2
SET industry = 'Crypto'
WHERE industry LIKE 'Crypto%';
```

```{sql Update_Country,connection=con,echo = TRUE}
SELECT DISTINCT country, TRIM(country, '.')
FROM layoffs_staging2
ORDER BY 1;
```

```{sql Update_US,connection=con,echo = TRUE}
UPDATE layoffs_staging2
SET country = RTRIM(country, '.')
WHERE country LIKE 'United States%';
```

```{sql View_Trimming Country,connection=con,echo = TRUE}
SELECT DISTINCT country, TRIM(country)
FROM layoffs_staging2
ORDER BY 1;
```

### Step 4: Formatting dates

```{r Date Formatting,echo = TRUE}
library(dplyr)
library(DBI)

# 1. Pull the data into R
df <- dbGetQuery(con, "SELECT * FROM layoffs_staging2")

# 2. Safely parse dates regardless of input format
if (is.numeric(df$date) || !any(is.na(as.numeric(na.omit(df$date))))) {
  # If dates are Excel-style numeric integers
  df$date <- as.Date(as.numeric(df$date), origin = "1970-01-01")
} else {
  # If dates are strings, try standard formats
  df$date <- as.Date(df$date, tryFormats = c("%Y-%m-%d", "%m/%d/%Y", "%m/%b/%Y", "%Y/%m/%d"))
}

# 3. Drop rows where dates absolutely couldn't be parsed to avoid downstream 'NA' strings
df <- df %>% filter(!is.na(date))

# 4. Convert to character string format 'YYYY-MM-DD' which SQLite expects
df$date <- as.character(df$date)

# 5. Push the cleaned, standardized table back to SQLite
dbWriteTable(con, "layoffs_staging2", df, overwrite = TRUE)
```

```{sql Check_FormattedDate,connection=con,echo = TRUE}
SELECT `date`
FROM layoffs_staging2
LIMIT 10;
```

```{sql Update_Table,connection=con,echo = TRUE}
UPDATE layoffs_staging2
SET company = TRIM(company);
```

```{sql Recheck_Table,connection=con,echo = TRUE}
SELECT `date`
FROM layoffs_staging2;
```

### Step 5: Dealing with Null or Blanks

```{sql Check for Null Values,connection=con,echo = TRUE}
SELECT *
FROM layoffs_staging2
WHERE total_laid_off IS NULL
AND percentage_laid_off IS NULL;
```

```{sql Update_Table_Industry,connection=con,echo = FALSE}
UPDATE layoffs_staging2
SET industry = NULL
WHERE industry = '';
```

```{sql Recheck_IndustryTable,connection=con,echo = TRUE}
SELECT *
FROM layoffs_staging2
WHERE industry is NULL
OR industry = '';
```

```{sql Outliers,connection=con,echo = TRUE}
SELECT *
FROM layoffs_staging2
WHERE company = '`Ballys Interactive`';
```

#### Step 5.1: FILLING MISSING VALUES in a FIELD based on KNOWN VALUES

```{sql Filling Missing Values,connection=con,echo = FALSE}
SELECT t1.industry, t2.industry
FROM layoffs_staging2 AS t1
JOIN layoffs_staging2 AS t2
	ON t1.company = t2.company
	AND t1.location = t2.location
WHERE (t1.industry IS NULL OR t1.industry = '')
  AND t2.industry IS NOT NULL;
```

```{sql Update Layoffs_Table,connection=con, echo = FALSE}
UPDATE layoffs_staging2 
SET industry = (
    SELECT t2.industry 
    FROM layoffs_staging2 t2
    WHERE t2.company = layoffs_staging2.company 
      AND t2.industry IS NOT NULL
    LIMIT 1
)
WHERE industry IS NULL OR industry = '';
```

```{sql RecheckTable_Layoffs,connection=con,echo = FALSE}
SELECT *
FROM layoffs_staging2;
```

### Step 6: Remove any columns

```{sql CheckColumns_LaidOff,connection=con,echo = FALSE}
SELECT *
FROM layoffs_staging2
WHERE total_laid_off IS NULL
AND percentage_laid_off IS NULL;
```

```{sql DeleteNULLValues,connection=con,echo = FALSE}
DELETE
FROM layoffs_staging2
WHERE total_laid_off IS NULL
AND percentage_laid_off IS NULL;
```

#### Step 6.1:DROPPING COLUMNS

```{sql DropColumns,connection=con,echo = TRUE}
ALTER TABLE layoffs_staging2
DROP COLUMN row_num;
```

```{sql RecheckTable3,connection=con,echo = FALSE}
SELECT * 
FROM layoffs_staging2;
```

## Exploratory Data Analysis of Company Layoffs (2020-2023)

```{sql EDA1,connection=con,echo = TRUE}
SELECT *
FROM layoffs_staging2;
```

```{sql EDA2,connection=con,echo = TRUE}
SELECT MAX(total_laid_off), MAX(percentage_laid_off)
FROM layoffs_staging2;
```

```{sql EDA3,connection=con,echo = TRUE}
SELECT company, SUM(total_laid_off)
FROM layoffs_staging2
GROUP BY company
ORDER BY 2 DESC;
```

```{sql EDA4,connection=con,echo = TRUE}
SELECT MIN(`date`), MAX(`date`)
FROM layoffs_staging2;
```

```{sql EDA5,connection=con,echo = FALSE}
SELECT country, SUM(total_laid_off)
FROM layoffs_staging2
GROUP BY country
ORDER BY 2 DESC;
```

```{sql EDA6,connection=con,echo = FALSE}
SELECT stage, SUM(total_laid_off)
FROM layoffs_staging2
GROUP BY stage
ORDER BY 2 DESC;
```

```{sql EDA7,connection=con,echo = FALSE}
SELECT country, SUM(total_laid_off)
FROM layoffs_staging2
GROUP BY country
ORDER BY 2 DESC;
```

### Rolling Totals Analysis

```{sql RollingTotalsofLayoffs,connection=con,output.var="rolling_layoffs",echo = TRUE}
WITH Rolling_Total AS (
  SELECT 
    substr(`date`, 1, 7) AS `MONTH`, 
    SUM(total_laid_off) AS Total_LaidOff
  FROM layoffs_staging2
  WHERE substr(`date`, 1, 7) IS NOT NULL
  GROUP BY `MONTH`
)
SELECT 
  `MONTH`, 
  Total_LaidOff,
  SUM(Total_LaidOff) OVER(ORDER BY `MONTH`) AS Rolling_total
FROM Rolling_Total
ORDER BY `MONTH` ASC;
```

```{r RollingTotalGraph, include=TRUE,echo=TRUE}
library(dplyr)
library(ggplot2)

# 1. Force Column Headers to Uppercase to prevent casing mismatch errors
colnames(rolling_layoffs) <- toupper(colnames(rolling_layoffs))

# 2. Convert YYYY-MM string paths into a true calendar date sequence
rolling_cleaned <- rolling_layoffs %>%
  filter(!is.na(MONTH), MONTH != "", MONTH != "NA") %>%
  mutate(
    # Appends day "-01" to "YYYY-MM" to enable clean chronologic timeline parsing
    Timeline_Date = as.Date(paste0(MONTH, "-01")),
    ROLLING_TOTAL  = as.numeric(ROLLING_TOTAL)
  ) %>%
  filter(!is.na(Timeline_Date))

# 3. Build the Corporate Portfolio Area-Line Visualization
RollingTotalsPlot <- ggplot(rolling_cleaned, aes(x = Timeline_Date, y = ROLLING_TOTAL)) +
  # Add a subtle, translucent professional canvas gradient shadow
  geom_area(fill = "#4A6572", alpha = 0.15) +
  # Main tracking growth line
  geom_line(color = "#34495E", linewidth = 1.2) +
  # Individual localized data nodes
  geom_point(color = "#E07A5F", size = 1.8) +
  
  # Format X-axis timeline ticks neatly spaced every 4 months
  scale_x_date(date_labels = "%b %Y", date_breaks = "4 months") +
  # Formats raw large numbers with readable commas (e.g., 150,000)
  scale_y_continuous(labels = scales::label_comma()) +
  
  labs(
    title = "Cumulative Company Layoffs",
    subtitle = "Chronological aggregation of global workforce restructuring volumes (2020 - 2023)",
    x = "Time",
    y = "Total Laidoff",
    caption = "Source: Real-world tech sector tracking metrics via Layoffs.fyi"
  ) +
  
  # APPLY SHARED STYLE HERE:
  professional_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Quick date text angle override

# 4. Save a high-fidelity vector copy asset to your local project directory
ggsave(
  filename = "Cumulative Rolling Total of Layoffs between 2020−2023.pdf",
  plot = RollingTotalsPlot,
  width = 10,
  height = 5.5,
  dpi = 300
)

# 5. Output cleanly onto your active notebook HTML markdown canvas
RollingTotalsPlot
```

### Company Layoffs Analysis
```{sql CompanyLayoffs1,connection=con,echo = TRUE}
SELECT TRIM(company) As Company, SUM(total_laid_off)
FROM layoffs_staging2
GROUP BY company
ORDER BY 2 DESC;
```

```{sql CompanyLayoffs2,connection=con,output.var="company_yearly_layoffs",echo = TRUE}
SELECT 
  company AS Company, 
  strftime('%Y', `date`) AS year_, 
  SUM(total_laid_off) AS Total_Laid_Off
FROM layoffs_staging2
WHERE `date` IS NOT NULL 
  AND `date` != '' 
  AND `date` != 'NA'
GROUP BY company, year_
HAVING year_ IS NOT NULL AND year_ != 'NA'
ORDER BY Total_Laid_Off DESC;
```


```{r YearlyCompanyLayoffsGraph,echo = TRUE}
library(dplyr)
library(ggplot2)

# 1. Standardize column names to UPPERCASE
colnames(company_yearly_layoffs) <- toupper(colnames(company_yearly_layoffs))

# 2. Find the top 10 companies using the new UPPERCASE column names
top_10_companies <- company_yearly_layoffs %>%
  group_by(COMPANY) %>%                                      
  summarise(GRAND_TOTAL = sum(TOTAL_LAID_OFF, na.rm = TRUE)) %>% 
  slice_max(GRAND_TOTAL, n = 10) %>%
  pull(COMPANY)                                              

# 3. Assign the pipeline outcome right from the start and pipe into ggplot
YearTrendLayoffs <- company_yearly_layoffs %>%
  filter(COMPANY %in% top_10_companies, !is.na(YEAR_)) %>%
  mutate(YEAR_ = as.factor(YEAR_)) %>% 
  ggplot(aes(x = reorder(COMPANY, TOTAL_LAID_OFF), y = TOTAL_LAID_OFF, fill = YEAR_)) +
    geom_col(position = "dodge") +
    coord_flip() +
    scale_fill_brewer(palette = "Dark2") + 
    labs(
      title = "Yearly Breakdown for Top 10 Laying-Off Companies (2020-2023)", 
      x = "Company", 
      y = "Number of Layoffs", 
      fill = "Year",
      caption = "Source: Real-world tech sector tracking metrics via Layoffs.fyi"
    ) +
  
    # APPLY SHARED STYLE HERE (Facets look great with the clean white canvas theme):
    scale_fill_manual(values = brand_colors) + 
    professional_theme +
    theme(panel.spacing = unit(1.5, "lines")) # Quick layout padding override

# 4. EXPORT THE PLOT IN HIGH QUALITY
ggsave(
  filename = "Yearly Trend of Company Layoffs 2020-2023.pdf",
  plot = YearTrendLayoffs,  # Explicitly tell ggsave which plot object to export
  #width = 10,
  #height = 6,
  #dpi = 300
)

# 5. Print the graph into the Markdown report
YearTrendLayoffs
```

### Layoffs Rankings

Using CAST

```{sql Top5Companies_20202023,connection=con,output.var="top_5_per_year",echo=TRUE}
WITH Company_Year AS (
  SELECT 
    company AS Company, 
    strftime('%Y', `date`) AS Years,
    SUM(total_laid_off) AS Total_Laid_Off
  FROM layoffs_staging2
  WHERE total_laid_off IS NOT NULL 
    AND `date` IS NOT NULL
    AND `date` != ''
    AND `date` != 'NA'
  GROUP BY company, Years
), 
Company_Year_Rank AS (
  SELECT 
    Company,
    Years,
    Total_Laid_Off,
    DENSE_RANK() OVER (PARTITION BY Years ORDER BY Total_Laid_Off DESC) AS Ranking
  FROM Company_Year
  WHERE Years IS NOT NULL 
    AND Years != 'NA' 
    AND CAST(Years AS INTEGER) >= 2020
)
SELECT Company, Years, Total_Laid_Off, Ranking
FROM Company_Year_Rank
WHERE Ranking <= 5;
```

```{r, include=FALSE, echo=FALSE}
print("--- DATA SUMMARY ---")
print(paste("Total rows fetched:", nrow(top_5_per_year)))

print("--- DATA STRUCTURE ---")
str(top_5_per_year)

print("--- FIRST 5 ROWS ---")
head(top_5_per_year, 5)
```

```{r Top5Companies_layoffs,echo = TRUE}
library(ggplot2)
library(dplyr)

# 1. FORCE CASE-INSENSITIVITY: Convert data frame headers to uppercase
colnames(top_5_per_year) <- toupper(colnames(top_5_per_year))

# 2. EMERGENCY SAFETY CHECK: Prevents the knit from breaking if SQL returns 0 rows
if (nrow(top_5_per_year) == 0) {
  
  plot(0, type='n', axes=FALSE, ann=FALSE)
  text(1, 0, "ERROR: SQL Query returned 0 rows. Check your date formatting.", cex=1.2, col="red")
  
} else {
  
# 3. RUN PLOTTING PIPELINE WITH UPDATED UPPERCASE COLUMN NAMES
top5companies_layoffs <- top_5_per_year %>%
  # Filter out empty rows just in case the dataset came back blank
  filter(!is.na(COMPANY), COMPANY != "") %>% 
  
  # Ensure YEARS is treated as a factor so reordering works cleanly
  mutate(YEARS = as.factor(YEARS)) %>% 
  mutate(COMPANY = tidytext::reorder_within(COMPANY, TOTAL_LAID_OFF, YEARS)) %>%
  
  ggplot(aes(x = COMPANY, y = TOTAL_LAID_OFF, fill = YEARS)) +
  geom_col(show.legend = FALSE) +
  tidytext::scale_x_reordered() +
  coord_flip() +
  facet_wrap(~YEARS, scales = "free_y") +
  
  scale_fill_brewer(palette = "Set2") + 
  
  labs(
    title = "Top 5 Companies with Highest Layoffs Per Year",
    subtitle = "(2020-2023)",
    x = "Company",
    y = "Total Laid Off",
    caption = "Source: Real-world tech sector tracking metrics via Layoffs.fyi"
  ) +
  
  # APPLY SHARED STYLE HERE (Facets look great with the clean white canvas theme):
    scale_fill_manual(values = brand_colors) + 
    professional_theme +
    theme(panel.spacing = unit(1.5, "lines")) # Quick layout padding override

# 3. EXPORT THE PLOT IN HIGH QUALITY
ggsave(
  filename = "Top 5 Companies Layoffs 2020-2023.pdf", # The name of your file
#  width = 10,                            # Width in inches
#  height = 6,                            # Height in inches
#  dpi = 300                              # 300 DPI is standard crisp print quality
)

# 3. Print the graph into the Markdown report
top5companies_layoffs
}
```

```{r reconnect-to-mariadb, message=FALSE, warning=FALSE,include=FALSE}
# 1. Disconnect from the temporary SQLite session
if (exists("con")) dbDisconnect(con)

# 2. Reconnect to your local MariaDB Parks & Rec instance
con <- dbConnect(RMariaDB::MariaDB(),
                 user = "root",
                 password = "Munna123@#",
                 host = "127.0.0.1",
                 port = 3306,
                 dbname = "parks_and_recreation")

# 3. Reset the knitr environment driver option back to MariaDB
knitr::opts_chunk$set(connection = "con")
```
